SPP-Perc: Peceptron Learning Based Enhanced Signature Path Prefetcher

### 2018-12-04 intro.tex
Changes in Design. Minor edits in intro / method / results
spell check, editing for grammar, clarity, style, put in some questions preceded by 'djimenez'

An ideal prefetching scheme would perfectly capture a program's memory access pattern, and then
predict and pre-load the needed data into the processor's caches in a timely manner. 
Memory access patterns may be simple, such as accessing every item in an array with a for-loop,
or very complex, such as chasing pointers through dynamically-allocated memory.
Practical prefetchers face challenges not only in predicting what data will be useful in the future,
but also in timing when data should be prefetched, deciding which level of the cache hierarchy
the data should be stored in, and deciding what data should be evicted from the caches to accomodate
the prefetched data.

@@

Figure~\ref{Fig:Motivation} illustrates the above scenario in a working manner.
Here we consider a state-of-the-art lookahead prefetcher -- SPP. We keep on 
re-tuning its thresholds, each time getting a variant with an increasing lookahead 
depth coverage. The figure depicts the behavior of the prefetcher variants running 
on the trace \textit{603.bwaves\_s-1740B}. The IPC, total number of prefetches 
suggested by the prefetcher (TOTAL\_PF) and the actual useful predictions (GOOD\_PF),
all have been normalized to the least aggressive version of the prefetcher. 
As the look-ahead depth of the prefetch suggestions increase, so do useful prefetches. 
But that comes at a cost of total prefetches inceasing at an even higher rate. This leads 
to cache pollution and bandwidth contention, as is clearly visible with the reduction in 
the IPC. Similar behaviour was observed in 12 other SPEC 2017 simpoints.

Therefore, a delicate balance between coverage and accuracy is required for a prefetcher to maximize
its performance impact.  Prefetchers are generally designed with internal mechanisms to
monitor their accuracy, with throttling thresholds that can be tuned for either coverage 
or accuracy.  The more irregular an application's memory access pattern is, the more difficult
it is to accurately predict every access, so a prefetcher will have to be tuned more toward
coverage (and away from accuracy) in order to gain any benefit.  This may be especially dangerous
to do in the context of a multi-core processor, where overly aggressive prefetching in one core can
waste shared resources, such as last-level cache capacity, and off-chip bandwidth~\cite{Friendly}.

@@

%For a prefetcher to capture highly irregular access patterns, it needs to be
%highly aggressive with high coverage. This causes the prefetcher to recommend many
%low accurracy prefetches that lead to the cache being polluted with useless data.  
%This effect is more pronounced in a multi-core scenario where the last-level cache is a
%shared resource~\cite{Friendly}.  Conversely, a conservative prefetcher would
%be highly accurate but might not make enough useful prefetches due to its lower coverage.

@@

In a single core configuration, PPF increases performance by 6.84\%
compared to the baseline prefetcher, SPP.  In a multi-core system running a
mixes of memory intensive SPEC CPU 2017 traces, PPF saw an improvement
of 11.9\% over SPP for a 4-core system, and 10.67\% for an 8-core system.

>>>>

An ideal prefetching scheme would perfectly capture a program's memory access pattern, and then
predict and pre-load the needed data into the processor's caches in a timely manner. 
Memory access patterns may be simple, such as accessing every item in an array with a for-loop,
or very complex, such as chasing pointers through dynamically-allocated memory.
Practical prefetchers face challenges not only in predicting what data will be useful in the future,
but also in timing when data should be prefetched, deciding which level of the cache hierarchy
the data should be stored in, and deciding what data should be evicted from the caches to accommodate
the prefetched data.

@@

Figure~\ref{Fig:Motivation} illustrates the above scenario in a working manner.
Here we consider a state-of-the-art lookahead prefetcher -- SPP. We keep on 
re-tuning its thresholds, each time getting a variant with an increasing lookahead 
depth coverage. The figure depicts the behavior of the prefetcher variants running 
on the trace \textit{603.bwaves\_s-1740B}. The IPC, total number of prefetches 
suggested by the prefetcher (TOTAL\_PF) and the actual useful predictions (GOOD\_PF),
all have been normalized to the least aggressive version of the prefetcher. 
As the look-ahead depth of the prefetch suggestions increase, so do useful prefetches. 
But that comes at a cost of total prefetches increasing at an even higher rate. This leads 
to cache pollution and bandwidth contention, as is clearly visible with the reduction in 
the IPC. Similar behavior was observed in 12 other SPEC CPU 2017 simpoints.

Therefore, a delicate balance between coverage and accuracy is required for a prefetcher to maximize
its performance impact.  Prefetchers are generally designed with internal mechanisms to
monitor their accuracy, with throttling thresholds that can be tuned for either coverage 
or accuracy.  The more irregular an application's memory access pattern is, the more difficult
it is to accurately predict every access, so a prefetcher will have to be tuned more toward
coverage (and away from accuracy) in order to gain any benefit.  This may be especially dangerous
to do in the context of a multi-core processor, where overly aggressive prefetching in one core can
waste shared resources, such as last-level cache (LLC) capacity, and off-chip bandwidth~\cite{Friendly}.

@@

%For a prefetcher to capture highly irregular access patterns, it needs to be
%highly aggressive with high coverage. This causes the prefetcher to recommend many
%low accuracy prefetches that lead to the cache being polluted with useless data.  
%This effect is more pronounced in a multi-core scenario where the LLC is a
%shared resource~\cite{Friendly}.  Conversely, a conservative prefetcher would
%be highly accurate but might not make enough useful prefetches due to its lower coverage.

@@

In a single core configuration, PPF increases performance by 6.84\%
compared to the underlying prefetcher, SPP.  In a multi-core system running a
mixes of memory intensive SPEC CPU 2017 traces, PPF saw an improvement
of 11.9\% over SPP for a 4-core system, and 10.67\% for an 8-core system.

### 2018-12-04 method.tex
Changes in Design. Minor edits in intro / method / results
spell check, editing for grammar, clarity, style, put in some questions preceded by 'djimenez'

\subsection{Performance Model}
\label{Method-Model}
For testing and comparing PPF, we used the ChampSim~\cite{Champsim} simulator.
ChampSim is an enhanced version of the framework that was used for the 2nd
Data Prefetch Championship (DPC-2)~\cite{DPC_2}. We model 1-core, 4-core, and
8-core out-of-order machines. The details of the configuration parameters are
summarized in Table~\ref{tab:Sim_params}.

@@

\subsection{Workloads}
\label{Method-Workloads}
% djimenez: don't say this is the first time SPEC 2017 has been used this way.
% i don't know if that's true, and in any event the value is small.

%This is the first time that SPEC 2017 benchmark suite~\cite{SPEC2017}
%has been used to characterize and measure the prefetch performance.
We use the SPEC 2017 benchmark suite~\cite{SPEC2017}.  We use all the 20
workloads available in the SPEC 2017 suite.  Using the SimPoint~\cite{SimPoint}
methodology, we identified 95 different program segments of 1 Billion
instructions each.

@@

\textit{Multi-core performance:} For multi-application workloads, we generate
100 random mixes and another 100 mixes from the memory intensive subset of
SPEC 2017.  For 4-core workload, 200 Million instructions are used for warm-up
and additional 1 Billion instruction simulated for collecting statistics.
Each CPU keeps executing its workload till the last CPU completes one billion
instructions after warm-up.  For collecting IPC and other data, only the first
billion instructions are considered as the region of interest.

@@

\textit{Validation:} We cross-validated our PPF model using SPEC
2006~\cite{SPEC2006} and CloudSuite~\cite{CloudSuite} benchmarks.  For
single-core SPEC 2006, we developed 94 simpoints spread across all the 29
applications. For multi-core, we followed the same methodology as SPEC 2017.
For CloudSuite, we used the traces made available for the 2nd Cache
Replacement Competition (CRC-2)~\cite{CRC_2}.  The traces include 4 4-core
applications with 6 distinct phases per application.

In total, we used 285 traces representing workloads across 53 applications.
Throughout the paper, we consider memory intensive subset as the
traces with LLC MPKI > 1.  This includes 48 out of the 95 simpoints
developed for SPEC 2017. For SPEC 2006  This includes 48 out of 94 simpoints.

\subsection{Preferchers Simulated}
\label{Method-Prefetchers}
We compared PPF against three of the latest state of the art hardware-only
prefetchers: Best Offset Prefetcher (BOP), DRAM Aware - Access Map Pattern
Matching (DA-AMPM)~\cite{DA_AMPM} and Signature Path Prefecher (SPP).  BOP was
the winner of 2nd Data Prefetching Championship.  DA-AMPM is the enhanced
version of AMPM, modified to account for DRAM row buffer locality.  SPP has
been shown to outperform BOP on SPEC 2006 traces.  For each of these, we
compare their speedups taking no prefetching as the baseline.

@@

\textbf{Global Pearson's Correlation}\newline For this experiment, we create a
dump of the state of perceptron weights at the end of all trace execution.
The weights obtained from running all the SPEC 2017 traces are concatenated.
Since the weights are collected at the end of individual trace execution, the
perceptron weights have attained a relatively stable value by now.  Hence,
this dump represents a `snap' of the trained perceptron weights across all the
SPEC 2017 traces.  The intuition here is that the feature with bulk of the
perceptron weights concentrated around 0 or small magnitude numbers show a
weak correlation with the prefetching outcome.  On the other hand, features
with most of the weights saturated around highest value (+15) show a high
positive correlation and the features with weights close to the lowest value
(-16) show a strong negative correlation.

@@

\textbf{Per Trace Correlation} \newline Another important way to look
at the perceptron features is to see how much their contribution
varies across the traces.  Here we give special attention to features
with low P-values in.  Figure~\ref{Pvals_Traces} shows the
variation of P-values for three features : \textit{PC XOR Delta},
\textit{Signature XOR Delta} and 
\textit{PC XOR Depth}; across all the SPEC 2017 traces.  
For simplicity, the traces are arranged in an
increasing order of contribution made by the feature and the trace
names have been omitted.  It can be seen that even features with a low
overall correlation provide useful correlation (magnitude > 0.5) for
a significant number of traces.

\textbf{Trimming Features Using Correlation} \newline Besides
providing interesting insights into prefetching behaviour, P-value can
also be used for feature selection and prefetcher tuning.

@@

Secondly, studying the relative importance of each feature enabled 
us to vary the number of entries dedicated for each feature.  
Features with higher Pearson's correlation, \textit{cache line} 
and \textit{page address}  were given most importance and
allowed full 12-bits of indexing.  Features like \textit{ip XOR delta} 
and \textit{ip XOR depth} with a low overall P-value 
were allocated fewer indexes in the feature table.

>>>>

\subsection{Performance Model}
\label{Method-Model}
For testing and comparing PPF, we use the ChampSim~\cite{Champsim} simulator.
ChampSim is an enhanced version of the framework that was used for the 2nd
Data Prefetch Championship (DPC-2)~\cite{DPC_2}. We model 1-core, 4-core, and
8-core out-of-order machines. The details of the configuration parameters are
summarized in Table~\ref{tab:Sim_params}.

@@

\subsection{Workloads}
\label{Method-Workloads}
% djimenez: don't say this is the first time SPEC CPU 2017 has been used this way.
% i don't know if that's true, and in any event the value is small.

%This is the first time that SPEC CPU 2017 benchmark suite~\cite{SPEC2017}
%has been used to characterize and measure the prefetch performance.
We use the SPEC CPU 2017 benchmark suite~\cite{SPEC2017}.  We use all the 20
workloads available in the SPEC CPU 2017 suite.  Using the SimPoint~\cite{SimPoint}
methodology, we identified 95 different program segments of 1 Billion
instructions each.

@@

\textit{Multi-core performance:} For multi-application workloads, we generate
100 random mixes and another 100 mixes from the memory intensive subset of
SPEC CPU 2017.  For 4-core workload, 200 Million instructions are used for warm-up
and additional 1 Billion instruction simulated for collecting statistics.
Each CPU keeps executing its workload till the last CPU completes one billion
instructions after warm-up.  For collecting IPC and other data, only the first
billion instructions are considered as the region of interest.

@@

\textit{Validation:} We cross-validated our PPF model using SPEC
CPU 2006~\cite{SPEC2006} and CloudSuite~\cite{CloudSuite} benchmarks.  For
single-core SPEC CPU 2006, we developed 94 simpoints spread across all the 29
applications. For multi-core, we followed the same methodology as SPEC CPU 2017.
For CloudSuite, we used the traces made available for the 2nd Cache
Replacement Competition (CRC-2)~\cite{CRC_2}.  The traces include 4 4-core
applications with 6 distinct phases per application.

In total, we used 285 traces representing workloads across 53 applications.
Throughout the paper, we consider memory intensive subset as the
traces with LLC MPKI > 1.  This includes 48 out of the 95 simpoints
developed for SPEC CPU 2017. For SPEC CPU 2006, this includes 48 out of 94 simpoints.

\subsection{Prefetchers Simulated}
\label{Method-Prefetchers}
We compared PPF against three of the latest state of the art hardware-only
prefetchers: Best Offset Prefetcher (BOP), DRAM Aware - Access Map Pattern
Matching (DA-AMPM)~\cite{DA_AMPM} and Signature Path Prefetcher (SPP).  BOP was
the winner of 2nd Data Prefetching Championship.  DA-AMPM is the enhanced
version of AMPM, modified to account for DRAM row buffer locality.  SPP has
been shown to outperform BOP on SPEC CPU 2006 traces.  For each of these, we
compare their speedups taking no prefetching as the baseline.

@@

\textbf{Global Pearson's Correlation}\newline For this experiment, we create a
dump of the state of perceptron weights at the end of all trace execution.
The weights obtained from running all the SPEC CPU 2017 traces are concatenated.
Since the weights are collected at the end of individual trace execution, the
perceptron weights have attained a relatively stable value by now.  Hence,
this dump represents a `snap' of the trained perceptron weights across all the
SPEC CPU 2017 traces.  The intuition here is that the feature with bulk of the
perceptron weights concentrated around 0 or small magnitude numbers show a
weak correlation with the prefetching outcome.  On the other hand, features
with most of the weights saturated around highest value (+15) show a high
positive correlation and the features with weights close to the lowest value
(-16) show a strong negative correlation.

@@

\textbf{Per Trace Correlation} \newline Another important way to look
at the perceptron features is to see how much their contribution
varies across the traces.  Here we give special attention to features
with low P-values in.  Figure~\ref{Pvals_Traces} shows the
variation of P-values for three features : \textit{PC XOR Delta},
\textit{Signature XOR Delta} and 
\textit{PC XOR Depth}; across all the SPEC CPU 2017 traces.  
For simplicity, the traces are arranged in an
increasing order of contribution made by the feature and the trace
names have been omitted.  It can be seen that even features with a low
overall correlation provide useful correlation (magnitude > 0.5) for
a significant number of traces.

\textbf{Trimming Features Using Correlation} \newline Besides
providing interesting insights into prefetching behavior, P-value can
also be used for feature selection and prefetcher tuning.

@@

Secondly, studying the relative importance of each feature enabled 
us to vary the number of entries dedicated for each feature.  
Features with higher Pearson's correlation, \textit{cache line} 
and \textit{page address}  were given most importance and
allowed full 12-bits of indexing.  Features like \textit{IP XOR delta} 
and \textit{IP XOR depth} with a low overall P-value 
were allocated fewer indexes in the feature table.

### 2018-12-04 related.tex
spell check, editing for grammar, clarity, style, put in some questions preceded by 'djimenez'

The idea of using concepts from machine learning to drive cache management mechanism has become widely popular. Perceptrons in microarchitecture were first introduced by Jimenez and Lin ~\cite{Perc_Branch} for branch prediction. Their predictor consists of a table of weights that was indexed by the program counter associated to the branch. The dot product of the corresponding weight and a history of branches is thresholded against a magnitude to make a prediction. The perceptron-based predictor helped differentiate which branch history outcomes were relevant for marking a prediction for a certain branch, which yield higher accuracy than previously proposed branch predictors. This reason motivated the use of perceptrons in other speculation mechanisms.

@@

%%MOVE IT SOMEWHERE
Peled \textit{et.al.} proposed using on-line reinforcement learnign to dynamically scale the maginitude of input information feed into their prefetcher mechanism ~\cite{Semantics}. The major drawback of this work is the need of compiler support. SVMs and logistic regression were used by Liao \textit{et.al.} to do an off-line optimal feature search for their prefetcher.

>>>>

The idea of using concepts from machine learning to drive cache management mechanism has become widely popular. Perceptrons in microarchitecture were first introduced by Jim\'enez and Lin ~\cite{Perc_Branch} for branch prediction. Their predictor consists of a table of weights that was indexed by the program counter associated to the branch. The dot product of the corresponding weight and a history of branches is thresholded against a magnitude to make a prediction. The perceptron-based predictor helped differentiate which branch history outcomes were relevant for marking a prediction for a certain branch, which yield higher accuracy than previously proposed branch predictors. This reason motivated the use of perceptrons in other speculation mechanisms.

@@

%%MOVE IT SOMEWHERE
Peled \textit{et.al.} proposed using on-line reinforcement learning to dynamically scale the magnitude of input information feed into their prefetcher mechanism ~\cite{Semantics}. The major drawback of this work is the need of compiler support. SVMs and logistic regression were used by Liao \textit{et.al.} to do an off-line optimal feature search for their prefetcher.

### 2018-12-04 results.tex
Completing Methodology, Images
Changes in Design. Minor edits in intro / method / results
spell check, editing for grammar, clarity, style, put in some questions preceded by 'djimenez'

\begin{figure}[h]
\includegraphics[width=\columnwidth]{SPEC2017}
\caption{SPEC 2017 Single-Core IPC Speedup}
\label{Fig:SPEC2017_1core}
\end{figure}

@@

This section discusses the results obtained from running PPF, in
terms of prefetch cache coverage and Instruction Per
Cycle (IPC).  For SPEC 2017 suite, first we present the
results for single-threaded workloads then for multi-core
workloads.

@@

Figure~\ref{Fig:SPEC2017_1core} shows the IPC speedup obtained by
PPF, compred to BOP, DA-AMPM and SPP. 
In the figure we have depicted selected traces, where PPF shows a 
noticeable improvement or shows some degradation, along with the 
geometric mean improvement on the memory intensive subset of SPEC 2017, 
and finally for the complete SPEC 2017 benchmark. 
All the results have been normalized to baseline \textit{i.e.}, no
prefetching.

As can be seen, PPF yields a speedup of \textbf{42.8\%} over
the baseline. This is equivalent to \textbf{7.8\%} over DA-AMPM, \textbf{9.7\%} over BOP
and \textbf{6.92\%} over SPP.  Out of the 95 simpoints developed for SPEC 2017, 
PPF nearly matches or outperforms most of the prefetchers on 91 simpoints.
Only in 607.cactuBSSN and 654.roms, PPF fails to match the improvement 
offered by BOP.

At its peak, PPF manages to get the IPC speedup of a factor of
over \textbf{3.39x} on the trace \textit{602.gcc\_s-2226B}.  This also
corresponds to speedup gain of \textbf{60\%} over the next best
prefetcher -- BOP. 
In general, benchmarks \textit{bwaves, mcf, xalancbmk} 
and \textit{fotonik3d} benefit the most from PPF, with the speedup over
SPP ranging from \textbf{10\% to 25\%}.

On the full SPEC 2017 suite, PPF improves the baseline by \textbf{20.9\%},
which is \textbf{3.44\%} more than the next best prefetcher -- SPP.

@@

\textbf{Coverage}\newline
A prefetcher coverage can be defined as the ratio of number of misses 
avoided through prefetching to the number of misses in presence of no 
prefetchig.
Figure~\ref{Fig:Coverage} shows the fraction of misses in L2 and Last
Level Caches avoided by the various prefetchers.
PPF has the highest coverage of all the prefetchers simulated. Using PPF 
on SPEC 2017 benchmark, 
misses can be reduced by \textbf{62.5\%} and \textbf{82.8\%} in L2 Cache and Last Level 
Cache respectively. For the same benchmark, the next best prefetcher -- 
DA-AMPM covers \textbf{48.6\%} and \textbf{72.8\%} of the misses respectively.

This superior coverage of PPF can be attributed to aggressive 
re-tuning of the underlying SPP. At the same time, the Perceptron Filter
makes sure than this high coverage does not lead to an increased 
cache pollution.

@@

\textit{4-core environment} Figure~\ref{Fig:4Core_SPEC2017} (a) 
shows a comparison of speedups
obtained on 4-core mix of memory intensive subset of SPEC 2017.  We
plot all the 4 prefetchers, normalized to the baseline.  The workloads
have been sorted in the increasing order of the speedup.
As can be seen,
PPF offers an IPC improvement by \textbf{51.7\%} on these traces.  
Compared to the base prefetcher -- SPP only, this is \textbf{11.9\%} more.
As compared to the next best prefetcher -- DA-AMPM, this is 
This is \textbf{10.2\%} and \textbf{17.4\%} over DA-AMPM, 
the next best prefetcher, and BOP, respectively.

On a different set of fully random SPEC 2017 4-core mixes, 
PPF provides an IPC speedup of 
\textbf{26.2}\% over baseline, which is \textbf{5.7}\% more than just SPP.

@@

\textit{8-core environment} The sorted comparison of speedup obtained 
on the memory intensive 8-core mixes is depicted in Fig~\ref{Fig:8Core_SPEC2017}.
PPF improves baseline performance by \textbf{38.6\%} and that is \textbf{10.7\%} 
more than using just SPP. 
For completely random set of SPEC 2017 mixes, PPF improves performance by 
\textbf{23.6\%} over the baseline, corresponding to \textbf{4.8\%} over SPP.
This increased improvement achieved by PPF over the base engine SPP,
in a multi-core environment is not surprising. 
Since PPF has an extremely intelligent filter, it ends up 
eliminating useless prefetches before they can create cache
pollution in the shared LLC.

As can be seen, BOP offer a better improvement than 
SPP, for the memory intensive mix, This can be attributed 
to BOP's inherent aggressive nature. DA-AMPM is also ahead of
SPP in both the mixes. Interestingly, in all these cases, 
PPF consistently outperforms the best performing prefetcher.

@@

Figure \ref{Fig:AddnConstr}(a) and (b) show the performance of PPF in a 
reduced LLC and low bandwidth constraints respectively. For the sake of brevity, 
we have only shownselected traces, followed
by the overall performance on the complete benchmark.
As can be seen, \textit{mcf} in low bandwidth conditions is prefetch averse.
In general, any prefetcher yields a negative speedup on that trace.
On \textit{roms} and \textit{cactuBSSN}, PPF is unable to match the performance 
achieved by the best prefetcher. On the other hand, PPF outperforms all the 
other prefetchers on \textit{xalancbmk} and \textit{imagick} benchmarks.
Overall, PPF gives a better improvement under small LLC condition and 
matches the best prefetcher -- BOP under low DRAM bandwidth condition.

\begin{figure}[ht]
\includegraphics[width=\columnwidth]{SPEC2006}
\caption{SPEC 2006 Single-Core IPC Speedup}
\label{Fig:SPEC2006_1core}
\end{figure}

@@

Here we try to demonstrate the robustness of PPF by testing it on
completely unseen benchmarks, namely SPEC 2006 and CloudSuite.

Fig~\ref{Fig:SPEC2006_1core} shows the speed-up achieved by BOP, SPP and PPF on
selected traces of SPEC 2006, along with the overall geometric mean
over whole of the benchmark and over memory-intensive subset of the
SPEC 2006 benchmark.  As can be seen, PPF provides a speed-up of
\textbf{46.1\%} over baseline, on the memory intensive subset of SPEC 2006
benchmark. This corresponds to \textbf{6.66\%} over SPP and \textbf{9.86\%} over BOP and DA-AMPM.  
On the whole of the SPEC 2006 suite, the speedup obtained is \textbf{22.4\%},
which is \textbf{3.34\%} ahead of SPP.

For 4-core memory intensive mixes, PPF improves the baseline by \textbf{59.2\%}, 
\textbf{8.7\%} ahead of SPP. For 8-core memory intensive mixes, the IPC speedup 
over baseline is \textbf{47.9\%}, \textbf{11.5\%} ahead of SPP.

Figure~\ref{Fig:CloudSuite} shows the performance benefit comparison of all 
the prefetch schemes on 4 different applications in the CloudSuite benchmark. 
In general, these applications are prefetch agnostic. Even in that scenario, 
PPF manages a \textbf{3.7\%} improvement over no prefetching, making it ahead of the 
next best prefetcher -- SPP, which provides a 3.08\% IPC speedup.

As can be seen, the performance is very consistent given no prior
parameter tuning was done keeping these benchmarks in mind.  We attribute
this to the inherent adaptability of the perceptron model.  In
general, perceptron weights are able to adjust in real-time so as to
find the best possible correlation between the output and the given set of 
features.

>>>>

\begin{figure}[ht]
\includegraphics[width=\columnwidth]{SPEC2017}
\caption{SPEC CPU 2017 Single-Core IPC Speedup}
\label{Fig:SPEC2017_1core}
\end{figure}

@@

This section discusses the results obtained from running PPF in terms of
prefetch cache coverage and speedup. For the SPEC CPU 2017 benchmarks, first
we present the results for single-threaded workloads then for multi-core
workloads.

@@

% djimenez: why is this figure described this way?  it's not OK to just
% highlight noticeable improvement. the benchmarks must be selected by some
% unbiased criteria, e.g. some minimum MPKI under LRU or some minimum speedup
% under a baseline prefetcher that isn't PPF.

Figure~\ref{Fig:SPEC2017_1core} shows the IPC speedup obtained by PPF,
compared to BOP, DA-AMPM and SPP.  In the figure we have depicted selected
traces where PPF shows a noticeable improvement or shows some degradation,
along with the geometric mean improvement on the memory intensive subset of
SPEC CPU 2017, and finally for the complete SPEC CPU 2017 benchmark.  All the
results have been normalized to the baseline of no prefetching.

% djimenez: you've inconsistently italicized and not italicized benchmark
% names. i'll help you out here; i like putting them in a typewriter font so
% i'll do that throughout this section. also, you've inconsistent put and not put
% the number of the benchmark. i'll fix that too.

PPF yields a speedup of \textbf{42.8\%} over the baseline. This is equivalent
to \textbf{7.8\%} over DA-AMPM, \textbf{9.7\%} over BOP and \textbf{6.92\%}
over SPP.  Out of the 95 simpoints developed for SPEC CPU 2017, PPF nearly
matches or outperforms most of the prefetchers on 91 simpoints.  PPF fails to
match the improvement offered by BOP only for {\tt 607.cactuBSSN\_s} and {\tt
654.roms\_s}.

At its peak, PPF manages a speedup of a factor of over \textbf{3.39x} on the
trace {\tt 602.gcc\_s-2226B}.  This also corresponds to speedup gain of
\textbf{60\%} over the next best prefetcher -- BOP.  In general, benchmarks
{\tt 603.bwaves\_s}, {\tt 605.mcf\_s}, {\tt 623.xalancbmk\_s} and {\tt
649.fotonik3d\_s} benefit the most from PPF, with the speedup over SPP ranging
from \textbf{10\% to 25\%}.

% djimenez: is this geometric mean? if so, say so.
On the full SPEC CPU 2017 suite, PPF improves the baseline by \textbf{20.9\%},
which is \textbf{3.44\%} better than the next best prefetcher -- SPP.

@@

% djimenez: this is a weird way to make a header for a section. use \paragraph or \subsection

\textbf{Coverage}\newline
Prefetcher coverage is defined as the ratio of the number of misses avoided
through prefetching to the number of misses with no prefetching.

Figure~\ref{Fig:Coverage} shows the fraction of misses in the L2 and LLC
avoided by the various prefetchers.  PPF has the highest coverage of all the
prefetchers simulated. On the SPEC CPU 2017 benchmarks, PPF reduces misses by
\textbf{62.5\%} and \textbf{82.8\%} in the L2 and LLC respectively. For the
same benchmark, the next best prefetcher, DA-AMPM, covers \textbf{48.6\%} and
\textbf{72.8\%} of the misses respectively.

This superior coverage of PPF can be attributed to aggressive re-tuning of the
underlying SPP, enabled by the Perceptron Filter making sure the high coverage
does not lead to increased cache pollution.

@@

\textit{4-core environment} Figure~\ref{Fig:4Core_SPEC2017}(a) shows a
comparison of speedups obtained on 4-core mixes of a memory intensive subset
SPEC CPU 2017.  We plot all 4 prefetchers, normalized to the baseline.  The
workloads have been sorted in increasing order of the speedup.  PPF offers a
speedup of \textbf{51.7\%} on these traces, and improvement of \textbf{11.9\%}
over the baseline SPP, \textbf{10.2\%} over the next DA-AMPM, and
\textbf{17.4\%} over BOP.

On a different set of fully random SPEC CPU 2017 4-core mixes, PPF provides an
IPC speedup of \textbf{26.2}\% over the baseline, which is an improvement of
\textbf{5.7}\% over SPP.

@@

\textit{8-core environment} The sorted comparison of speedups on the memory
intensive 8-core mixes is shown in Figure~\ref{Fig:8Core_SPEC2017}.  PPF
improves baseline performance by \textbf{38.6\%}, an improvement of
\textbf{10.7\%} over SPP.  For a random set of SPEC CPU 2017 mixes, PPF
improves performance by \textbf{23.6\%} over the baseline, corresponding to
\textbf{4.8\%} over SPP.  This increased improvement achieved by PPF over the
base engine SPP in a multi-core environment is expected as PPF has a very
accurate filter, it eliminates useless prefetches before they can cause
pollution in the shared LLC.

BOP offers a better improvement than SPP for the memory intensive mixes. This
superiority can be attributed to BOP's inherent aggressive nature.  DA-AMPM is
also ahead of SPP in both the mixes. Interestingly, in all these cases, PPF
consistently outperforms the best performing prefetcher.

@@

Figure~\ref{Fig:AddnConstr}(a) and (b) show the performance of PPF in a
reduced LLC and with low bandwidth constraints, respectively. For the sake of
brevity, we have only shown selected traces followed by the overall
performance on the complete benchmark.  Benchmark {\tt 605.mcf\_s} in low
bandwidth conditions is prefetch averse.  In general, any prefetcher yields a
negative speedup on that trace.  On {\tt 654.roms\_s} and {\tt
607.cactuBSSN\_s}, PPF is unable to match the performance achieved by the best
prefetcher. On the other hand, PPF outperforms all the other prefetchers on
{\tt 623.xalancbmk\_s} and {\tt 638.imagick\_s} benchmarks.  Overall, PPF gives
a better improvement under small LLC condition and matches the best
prefetcher, BOP, under low DRAM bandwidth conditions.

\begin{figure}[ht]
\includegraphics[width=\columnwidth]{SPEC2006}
\caption{SPEC CPU 2006 Single-Core IPC Speedup}
\label{Fig:SPEC2006_1core}
\end{figure}

@@

% djimenez: cite cloudsuite here? do we have a reference?

Here we demonstrate the robustness of PPF by testing it on different
benchmarks: SPEC CPU 2006 and CloudSuite.

% djimenez: is the improvement over BOP and DA-AMPM exactly the same 9.86%
% here? that's weird.

Figure~\ref{Fig:SPEC2006_1core} shows the speed-up achieved by BOP, SPP and
PPF on traces from SPEC CPU 2006, along with the overall geometric mean over
the whole suite and over a memory-intensive subset.  PPF provides a speedup of
\textbf{46.1\%} over the baseline on the memory intensive subset of SPEC CPU
2006 benchmark, giving an improvement of \textbf{6.66\%} over SPP and
\textbf{9.86\%} over BOP and DA-AMPM.  On the whole of the SPEC CPU 2006
suite, the speedup is \textbf{22.4\%}, an improvement of \textbf{3.34\%} over
SPP.

For 4-core memory intensive mixes, PPF improves the baseline by
\textbf{59.2\%}, \textbf{8.7\%} ahead of SPP. For 8-core memory intensive
mixes, the speedup over the baseline is \textbf{47.9\%}, \textbf{11.5\%} ahead
of SPP.

Figure~\ref{Fig:CloudSuite} shows the performance benefit comparison of all
the prefetch schemes on 4 different applications in the CloudSuite benchmark.
In general, these applications are prefetch agnostic. Even so, PPF manages a
\textbf{3.7\%} improvement over no prefetching, putting it ahead of the next
best prefetcher, SPP, which provides a 3.08\% speedup.

We developed PPF to yield good performance on the SPEC CPU 2017 benchmarks.
Nevertheless, the performance is consistently good on other benchmark suites.
We attribute this fact to the inherent adaptability of the perceptron model.
In general, perceptron weights are able to adjust in real-time so as to find
the best possible correlation between the output and the given set of
features.
